---
title: "PA1_template"
author: "Shu Guo"
date: "Tuesday, August 14, 2014"
output: html_document
---

# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r, echo=TRUE}
dat <- read.csv("activity.csv", header = T, 
        colClasses = c("numeric", "character", "numeric"))
# Check the total observations:
nrow(dat) == 17568
# Change the varialbe date from character to "Date" class.
dat$date <- as.Date(dat$date, format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

1. Make a histogram of the total number of steps taken each day

2. Calculate and report the mean and median total number of steps taken per day

```{r, echo=TRUE}
# First delete missing values.
datq1 <- na.omit(dat)
# Create a factor variable for each day
datq1$dfactor <- factor(datq1$date)
# create a new data set for the total number of steps taken each day.
s <- as.data.frame(tapply(datq1$steps, datq1$dfactor, sum))
s$date <- row.names(s)
names(s) <- "steps"
hist(s$steps, main = "Total Number of Steps Taken Each Day", 
     col = "blue", xlab = "Steps")
meanSteps <- format(mean(s$steps), scientific = FALSE, digits = 7)
medianSteps <- format(median(s$steps), scientific = FALSE)
```

The mean of total number of steps taken per day is `r meanSteps`, and the median of total number of steps taken per day is `r medianSteps`.


## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=TRUE}
# Create a new data frame for this question
s2 <- datq1
# Create a factor variable for intervals
s2$finterval <- factor(s2$interval)
# Create a data frame for steps average across all days.
s2 <- as.data.frame(tapply(s2$steps, s2$finterval, mean))
names(s2) <- "meanSteps"
s2$interval <- row.names(s2)
# Make the plot using ggplot2
library(ggplot2)
ggplot(s2, aes(x = as.numeric(interval), y = meanSteps, group = 1)) + 
    geom_line() +
    xlab("5-minute interval") +
    ylab("average steps taken across all days") +
    ggtitle("Average Daily Activity Pattern") +
    scale_x_discrete(breaks=c(500, 1000, 1500, 2000, 2500))
MaxSteps <- max(s2$meanSteps)
MaxStepsInt <- s2$interval[which(s2$meanSteps == MaxSteps)]
```

Therefore, the 5-minute interval that contains the maximum number of steps is `r MaxStepsInt`

## Imputing missing values
```{r, echo = TRUE}
# Calculate the number of row with NA in steps variable
NAs <- nrow(dat[dat$steps == NA, ])
# Now we can fill all the missing values using average steps
# acrossing all days obtained in previous part.
# First create a new data set with mean steps acrossing all days.
newdat <- cbind(dat, rep(s2$meanSteps, (nrow(dat)/nrow(s2))))
names(newdat)[4] <- "meanSteps"
# fill missing values
for (i in 1:nrow(newdat)) {
    newdat$steps[i] <- (ifelse(is.na(newdat$steps[i]), 
    newdat$meanSteps[i], newdat$steps[i]))}
# Create a factor variable to calculate sum of steps each day
newdat$fdate <- factor(newdat$date)
# Now calculate the total numbers of steps take each day
SumSteps <- aggregate(newdat$steps, list(newdat$fdate), sum)
names(SumSteps) <- c("Date", "Steps")
# Make the histogram
hist(SumSteps$Steps, main = "Total Number of Steps Taken Each Day", 
     col = "green", xlab = "steps")
# Calculate the mean and median total number of steps taken per day
meanSteps2 <- format(mean(SumSteps$Steps), scientific = FALSE)
medianSteps2 <- format(median(SumSteps$Steps), scientific = FALSE)
```

The total number of rows with NAs is `r NAs`. The mean and median for the new dataset is `r meanSteps2` and `r medianSteps2`, respectively.
These values are very close to the estimates from the first part of the assignment.


## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE}
v <- newdat[, 1:3]
v$weekday <- weekdays(v$date)
for (i in 1:nrow(v)) {
    if (v$weekday[i] == "Saturday" || v$weekday[i] == "Sunday"){
        v$fweekday[i] <- "weekend"}
    else v$fweekday[i] <- "weekday"
}
v$fweekday <- factor(v$fweekday, levels = c("weekend", "weekday"))
gv <- aggregate(v$steps, list(v$interval, v$fweekday), mean)
names(gv) <- c("interval", "weekday", "steps")
# Make the graph using ggplot2
library(ggplot2)
ggplot(gv, aes(x = interval, y = steps)) + geom_line() +
    facet_wrap(~ weekday, ncol = 1) + ylab("Number of steps") + 
    xlab("Interval") +
    theme(strip.text = element_text(face="bold", size=rel(1.2)),
          strip.background = element_rect(fill="lightblue", 
                        colour="black", size=1))
```

From the two plots we can the patterns of steps taken are very similar between weekends and weekdays. But we still can see some interesting difference between the two periods: there are more steps taken in weekdays among the intervals from 500 to 1000, and people take more steps during weekends within intervals from 1000 to 1500.
